% =====================================================================
%  WORKING PAPER SKELETON  —  read-and-iterate draft
%  Compile: pdflatex main.tex  (plain article class; swap to cvpr.sty later)
% ---------------------------------------------------------------------
%  MAIN IDEA (plain words):
%    When you cut / break / slice open a 3D Gaussian-Splat object, the
%    inside it shows is FABRICATED — a splat is a hollow shell. Every
%    existing method invents that interior with a LEARNED / OPTIMIZED
%    prior: PhysTalk deforms only the surface (no interior at all);
%    FruitNinja hallucinates interior TEXTURE with per-object diffusion
%    optimization; GaussianFluent densifies interior gaussians with a
%    diffusion prior in the loop. All are slow, category-blind, and
%    per-object.
%    OUR INSIGHT: a VLM already KNOWS what is inside a watermelon, a
%    tree log, an onion. That knowledge is a TRAIN-FREE, STRUCTURAL
%    prior. The agent recognizes the object, reasons about its internal
%    layout, and WRITES A SHORT PROGRAM that places actual 3D structure
%    (radial seeds, concentric rings, nested layers) in the gaussians —
%    which a topology-changing op then reveals. No training, no
%    per-object optimization, no diffusion in the loop. Instant, cut-
%    consistent, and — because it is instant — it SCALES interiors from
%    one object to a whole scene (the enabling piece for explorable,
%    changeable worlds).
%    THE BET (single sentence): if our interior does not look OBVIOUSLY
%    more STRUCTURALLY CORRECT than FruitNinja's texture fill, there is
%    no paper. Novelty is narrow and lives entirely in
%    "world-knowledge instead of a learned prior, STRUCTURE not TEXTURE."
%
%  TITLE OPTIONS (pick one):
%    - Structure, Not Texture: Train-Free, World-Knowledge-Grounded
%      Interiors for Gaussian Splatting
%    - The Agent Knows What's Inside: Train-Free Interior Authoring for
%      Gaussian-Splat Objects
%    - Cut It Open: World-Knowledge Interiors for Editable Gaussian
%      Splats, Without Training
%    - Inside Out: VLM-Authored Structural Interiors for Topology-
%      Changing Edits of 3D Gaussians
% =====================================================================
\documentclass[10pt,twocolumn,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}

\newcommand{\sysname}{\textsc{InsideOut}} % working codename — placeholder
\newcommand{\todo}[1]{{\color{red}\small[TODO: #1]}}
\newcommand{\note}[1]{{\color{teal}\small(#1)}}

\title{\vspace{-1em}\bfseries Structure, Not Texture:\\
Train-Free, World-Knowledge-Grounded Interiors\\ for Gaussian Splatting}
\author{Anonymous \and \todo{author list}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
When a 3D Gaussian-Splatting object is cut, broken, or sliced open, what
does its interior look like? A splat is a hollow shell of surface
primitives, so every existing answer \emph{fabricates} the inside.
PhysGaussian and PhysTalk deform only the surface and never expose an
interior; FruitNinja hallucinates interior \emph{texture} with a
per-object diffusion optimization; GaussianFluent densifies interior
gaussians with a diffusion prior in the loop. All of these priors are
\emph{learned or optimized}, blind to the identity of the specific
object, and slow. We observe that a vision--language model already
\emph{knows} what is inside a watermelon, a tree log, or an onion, and
that this knowledge is a \emph{train-free, structural} prior. We present
\sysname{}, a framework in which a VLM agent recognizes an object,
reasons about its internal layout, and \emph{writes a short program}
that instantiates that structure directly in the gaussians---radial
seeds, concentric rings, nested layers---which a topology-changing
operation then reveals. There is no training, no per-object
optimization, and no diffusion in the loop: interiors are authored in
seconds and stay consistent under arbitrary cut directions. Because it
is instant, the method scales editable, physically-changeable interiors
from single objects to whole scenes---a step toward explorable worlds
whose objects can be opened, cut, and broken. We demonstrate
structurally-correct interiors across \todo{N} objects and \todo{M}
operations, head-to-head against FruitNinja (structure vs.\ texture,
zero-optimization runtime), GaussianFluent (no diffusion-in-loop), and
PhysTalk (interior vs.\ none). We are explicit about the failure mode we
inherit: when an object's interior is genuinely unidentifiable, the
agent should \emph{abstain} rather than confabulate.
\end{abstract}

% =====================================================================
\section{Introduction}
\label{sec:intro}
\note{Arc: (1) worlds from media should be changeable, not just
viewable; (2) changing an object means exposing its inside; (3) splats
are hollow shells, so the inside is always invented; (4) prior inventions
are learned/optimized/slow/category-blind; (5) the VLM already knows the
inside --- use it as a train-free structural prior; (6) contributions.}

\todo{one motivating figure: same watermelon splat cut open ---
PhysTalk (deflates, no interior) $\mid$ FruitNinja-style texture fill
(smeared, wrong seed placement, cut-inconsistent) $\mid$ ours (radial
seeds placed by world knowledge, consistent under any cut).}

\paragraph{Contributions.}
\begin{itemize}\itemsep0.1em
  \item We reframe an object's interior as recoverable from a VLM's
    \emph{world knowledge}, used as a \emph{train-free structural prior}
    --- in place of the learned, optimized, or diffusion priors of prior
    work.
  \item An \emph{agentic, train-free} pipeline: recognize $\rightarrow$
    reason about internal layout $\rightarrow$ author a program over an
    interior/physics op-library $\rightarrow$ instantiate structured
    interior gaussians $\rightarrow$ reveal via a topology-changing
    operation. No per-object optimization, no diffusion in the loop.
  \item \emph{Structure, not texture}: explicit 3D internal geometry
    (seeds, rings, layers) that is \emph{consistent under any cut
    direction by construction}, unlike per-cross-section inpainting.
  \item A demonstration across objects and operations with direct
    head-to-heads against FruitNinja, GaussianFluent, and PhysTalk, and
    an honest treatment of \emph{unidentifiable} interiors (abstention),
    connecting to identifiability in single-view 3D generation.
\end{itemize}

% =====================================================================
\section{Related Work}
\label{sec:related}
\note{Position surgically --- the gap is narrow; each neighbor is one
axis away. State plainly what each does and does NOT do.}

\paragraph{Physics on Gaussian Splats.} PhysGaussian, PhysSplat, and
\textbf{PhysTalk} embed 3DGS in a physics simulator (MPM / Genesis) and
deform the visible surface. PhysTalk is the closest on the
\emph{agentic, train-free} axis---an LLM writes physics code---but it
never instantiates or reveals an interior; it transfers particle motion
back to existing surface gaussians. \todo{cite; confirm no
topology-change / interior}.

\paragraph{Interior generation for splats.} \textbf{FruitNinja}
(CVPR'25) first revealed a splat interior under slicing, via
\emph{per-object} progressive 2D-diffusion inpainting on cross-sections
plus voxel smoothing. It is not agentic, its prior is category-blind,
and its result is \emph{texture} with no guarantee that seeds/rings sit
where they should, nor that different cuts agree. \textbf{GaussianFluent}
(ICLR'26) densifies interior gaussians for brittle \emph{fracture}, but
with a \emph{diffusion prior in the loop}, not agentic and not
identity-grounded. \todo{cite both; this is the head-to-head axis.}

\paragraph{Inferring interior structure.} \textbf{TopoGaussian} infers
an object's \emph{load-bearing} interior topology by optimizing a
representation so a simulator explains observed motion---per-object
optimization, not semantic (``what mechanically works'', not ``this is a
watermelon''). \todo{cite.}

\paragraph{Agentic program authoring for 3D.} \textbf{Articraft} (2026)
has an LLM agent write a program against a DSL to build articulated 3D
\emph{meshes}. It establishes ``agent-writes-a-program-for-3D'' as a
paradigm---so that framing alone is not our novelty---but it does not
touch captured splats, interiors, or topology-changing reveals.
\todo{cite; use to pre-empt ``just Articraft for splats''.}

\paragraph{Editable / language-driven splats.} \todo{REALM (MLLM-agent
surface editing of 3DGS, no interior), VR-GS (tet-proxy interactive
physics), etc.}

% =====================================================================
\section{Method}
\label{sec:method}
\note{Reuse the phystalk substrate: load/render, solidify (axis-stab
interior fill), op-library (shatter/slice/break), agent-authoring core.
The NEW part is the structured interior authored from a VLM schema.}

\subsection{Overview}
Given a captured 3DGS object and a natural-language instruction
(``cut it open'', ``smash it''), a VLM agent produces (i) an
\emph{interior schema} describing the object's internal structure and
(ii) a \emph{transform program} over an op-library. The program
instantiates the structured interior as gaussians and applies a
topology-changing operation that reveals it. \todo{pipeline figure.}

\subsection{World knowledge as a structural prior}
The agent recognizes the object and emits a compact \emph{interior
schema}---a small typed program, e.g.
\texttt{radial\_seeds(count, size, region)},
\texttt{concentric\_rings(spacing, colors)},
\texttt{layers(thicknesses, colors)}, \texttt{solid\_core(color)},
\texttt{hollow(shell\_thickness)}. \note{The schema is the train-free
prior: it encodes what the VLM knows, and is category-correct by
construction rather than by inpainting luck.} \todo{define the schema
grammar / op-library table.}

\subsection{Instantiating structured interior gaussians}
We fill the hollow shell with interior gaussians (axis-stabbing
occupancy test, cf.\ \S\ref{sec:method}) and \emph{place structure}
according to the schema---seeds as discrete clusters at radial
positions, rings as concentric shells, layers as nested strata---rather
than copying surface color inward. \todo{algorithm box; contrast with
FruitNinja's OpaqueAtom texture fill.}

\subsection{Topology-changing operations}
\todo{slice / fracture / break / bite / drill from the op-library; the
reveal; two-sided consistency for fracture.}

% =====================================================================
\section{Experiments}
\label{sec:exp}
\note{Framed as a CAPABILITY DEMONSTRATION, not a benchmark. NO
ground-truth interior exists (FruitNinja says so) --- we do NOT chase
faithfulness-to-truth. Success = structure beats texture, train-free.}

\subsection{Success criteria}
\begin{itemize}\itemsep0.1em
  \item \textbf{Structural correctness} (primary): do placed elements
    match category-expected layout (seeds radial, rings concentric,
    layers nested)? Blind human + VLM-judge \note{randomized order,
    report both directions --- VLM judges have position bias}, plus a
    geometric check (discrete 3D seeds at plausible positions vs.\
    smeared texture).
  \item \textbf{Cut-consistency}: same object, multiple cuts $\rightarrow$
    interior stays 3D-consistent (no GT needed).
  \item \textbf{Zero-optimization runtime}: seconds vs.\ minutes --- the
    one unarguable number.
\end{itemize}

\subsection{Dataset}
\todo{$\sim$30--40 objects via TripoSplat / image$\rightarrow$3DGS,
weighted to visibly-structured interiors: watermelon (radial seeds),
tree log (rings), onion/cabbage (layers), pomegranate (arils), orange
(segments), cake (strata); a few real captured objects; plus
deliberately \emph{ambiguous} interiors (sealed gadget, random rock) for
the abstention/failure section. Category-reference cut photos for
judging --- NOT per-instance GT.}

\subsection{Head-to-heads}
\todo{vs FruitNinja (structure vs.\ texture + runtime); vs
GaussianFluent (no diffusion-in-loop, fracture); vs PhysTalk (interior
vs.\ none). Skipping any $\rightarrow$ reviewer reduces us to that method.}

% =====================================================================
\section{Limitations}
\label{sec:limits}
When an object's interior is genuinely \emph{unidentifiable} from its
exterior (a sealed device, an arbitrary rock), no prior---VLM or
otherwise---can recover it; the agent should \emph{abstain} rather than
confabulate. \note{This is where the unidentifiability thesis of our
prior work reappears --- as an honest limitation, not the paper's
thrust.} Our interiors are \emph{plausible and structurally category-
correct}, not faithful to the specific instance; and the VLM's knowledge
has gaps for rare or novel objects. \todo{expand.}

% =====================================================================
\section{Conclusion}
\label{sec:conc}
\todo{One instant, train-free agent authors structured interiors from
world knowledge --- the enabling piece for scaling changeable objects
from one to a whole explorable world.}

% =====================================================================
{\small
\bibliographystyle{plain}
% \bibliography{refs}   % \todo{add refs.bib: phystalk, fruitninja,
%   gaussianfluent, topogaussian, articraft, physgaussian, physsplat,
%   vr-gs, realm, physx-3d}
}
\end{document}
